---
title: 'Most Harmful Types of Severe Weather Events Across the United States'
author: "pszjmb1"
date: "Sunday, June 21, 2015"
output: html_document
---

# Synopsis

It is paramount to the public health and economy to optimise the use of resources in order to prevent and/or mitigate the effects of severe weather events. In order to help prioritise resource use, this report examines the [US National Weather Service Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) to determine the types of events most harmful to population health and the ones that have the greatest economic consequences. Our data analysis indicates that X events are most harmful to population health and Y events have the greatest economic consequences.

# Data Processing

## Computing Environment
```{r sessionInfo}
sessionInfo()
```

## Utility Functions
```{r utility}
# Load libraries
library("dplyr")     # Used for data handling
library("gdata")     # Used for data handling
library("ggplot2")   # Used for plotting
library("lubridate")   # Used for date handling
library("stringr")   # String manipulation 


timeprint <- function(s){
  # Prints out the given string with a datestamp
  #
  # args:
  #  s: The string to print
  s <- paste(format(Sys.time(), "%a %b %d %H:%M:%S %Y"),s,sep="> ")
  print(s)
}
```
## Data Source
The data for this report come from the U.S. National Oceanic and Atmospheric Administration's (NOAA). This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. Please see [the data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) for data integrity and providence information.
## Fetching and Loading Data
```{r getData, cache=TRUE}
if(!file.exists("data")){dir.create("data")}

## Download data
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("./data/FStormData.csv.bz2")){
    timeprint("Downloading data...")
    download.file(fileURL, destfile = './data/FStormData.csv.bz2', method="curl")
    outputMsg <- paste("Raw data downloaded from:", fileURL," ", date())    
    write(outputMsg, file = paste("./data/", "FStormData.csv.bz2", ".download.info.md", sep=""))
} else{
  con <- file("./data/FStormData.csv.bz2.download.info.md", "rt")  
  outputMsg <- readLines(con, 1)
}
timeprint(outputMsg)
stormData <- read.csv2("./data/FStormData.csv.bz2", sep=",", header=TRUE)
timeprint("Data loaded successfully.")

# Ensure that Variables are lowercase to ease data handling
names(stormData) <- tolower(names(stormData))

# Initial examination of the data
str(stormData)
```

## Data Exploration and Selection
```{r exploreData, cache=TRUE}
dim(stormData)
names(stormData)
stormData <- tbl_df(stormData)
stormData
```
```{r exploreData2, cache=TRUE}
head(stormData)
tail(stormData)
seed <- 314
set.seed(seed)
stormData[sample(nrow(stormData), 6),]
summary(stormData)
# Ensure correct the variable formats
sapply(stormData, class)    ## Get the data type of each variable
```
```{r exploreData3_ignoreRows, cache=TRUE}
# The assignment instructions warn: "More recent years should be considered more complete." Therefore we should exclide non-recent year rows.
## First we convert the date factor Variable into a date for easy processing 
stormData$bgn_date <- as.Date(stormData$bgn_date, format="%m/%d/%Y %H:%M:%S")
head(stormData$bgn_date)   # confirm date change
## Check weather events by year to visually inspect if there is a rise events which may correspond to when the dataset is more complete.
stormsPerYear <- stormData %>%
  mutate("year" = year(bgn_date)) %>%  group_by(year) %>% summarise(count = n())
ggplot(stormsPerYear, aes(year, count)) + geom_line() + ylab("Num events in dataset") + ggtitle("Num weather events per year")

# Results suggest a sharp rise after 1994, therefore exclude data from it and before
excludeBefore <- 1994
stormData <- stormData %>% filter(year(bgn_date) > 1994, fatalities > 0 | injuries > 0 | propdmg > 0 | cropdmg > 0)
dim(stormData)
```

```{r exploreData4_ignoreCols, cache=TRUE}
# Only include relevant Variables
stormData <- select(stormData, evtype, fatalities:cropdmgexp)
dim(stormData)
```

```{r exploreData5_evType, cache=TRUE}
# Examine the weather event types to ensure there are no anomalies
# unique(stormData$evtype)  # comment this in to do visual inspection
# Visual inspection shows many similar terms (such as HEAVY RAINS and HEAVY RAIN) and irrelevant values (such as Summary of March 14). Such data should be corrected:

# Exclude summaries
stormData <- subset(stormData, !grepl('summary', evtype, ignore.case = TRUE) )

# Group common terms
stormData$evtype <- str_trim(stormData$evtype)
stormData$evtype[grep("THUN", stormData$evtype, ignore.case = TRUE)] <- "THUNDERSTORM"
stormData$evtype[grep("CLOUD", stormData$evtype, ignore.case = TRUE)] <- "CLOUD"
stormData$evtype[grep("RAIN", stormData$evtype, ignore.case = TRUE)] <- "RAIN"
stormData$evtype[grep("HURRICANE", stormData$evtype, ignore.case = TRUE)] <- "HURRICANE"
stormData$evtype[grep("SNOW", stormData$evtype, ignore.case = TRUE)] <- "SNOW"
stormData$evtype[grep("HAIL", stormData$evtype, ignore.case = TRUE)] <- "HAIL"
stormData$evtype[grep("COLD", stormData$evtype, ignore.case = TRUE)] <- "COLD"
stormData$evtype[grep("CURRENT", stormData$evtype, ignore.case = TRUE)] <- "RIP CURRENT"
stormData$evtype[grep("TORNADO", stormData$evtype, ignore.case = TRUE)] <- "TORNADO"
stormData$evtype[grep("FLOOD", stormData$evtype, ignore.case = TRUE)] <- "FLOOD"
stormData$evtype[grep("WIND", stormData$evtype, ignore.case = TRUE)] <- "WIND"
stormData$evtype[grep("BLIZZARD", stormData$evtype, ignore.case = TRUE)] <- "BLIZZARD"
stormData$evtype[grep("WATERSPOUT", stormData$evtype, ignore.case = TRUE)] <- "WATERSPOUT"
stormData$evtype[grep("FOG", stormData$evtype, ignore.case = TRUE)] <- "FOG"
stormData$evtype[grep("HIGH TEMPERATURE", stormData$evtype, ignore.case = TRUE)] <- "HEAT"
stormData$evtype[grep("LOW TEMPERATURE", stormData$evtype, ignore.case = TRUE)] <- "COLD"
stormData$evtype[grep("MARINE", stormData$evtype, ignore.case = TRUE)] <- "MARINE MISHAP"
stormData$evtype[grep("HEAT", stormData$evtype, ignore.case = TRUE)] <- "HEAT"
stormData$evtype[grep("FIRE", stormData$evtype, ignore.case = TRUE)] <- "FIRE"
stormData$evtype[grep("STORM", stormData$evtype, ignore.case = TRUE)] <- "STORM"
stormData$evtype[grep("SLEET", stormData$evtype, ignore.case = TRUE)] <- "SLEET"
stormData$evtype[grep("ICE", stormData$evtype, ignore.case = TRUE)] <- "ICE"
stormData$evtype[grep("MICROBURST", stormData$evtype, ignore.case = TRUE)] <- "MICROBURST"
stormData$evtype[grep("FREEZ", stormData$evtype, ignore.case = TRUE)] <- "COLD"
stormData$evtype[grep("WARMTH", stormData$evtype, ignore.case = TRUE)] <- "HEAT"
stormData$evtype[grep("DRY", stormData$evtype, ignore.case = TRUE)] <- "DRY"
stormData$evtype[grep("RECORD HIGH", stormData$evtype, ignore.case = TRUE)] <- "HEAT"
stormData$evtype[grep("RECORD LOW", stormData$evtype, ignore.case = TRUE)] <- "COLD"
stormData$evtype[grep("GUSTNADO", stormData$evtype, ignore.case = TRUE)] <- "TORNADO"
stormData$evtype[grep("WINTRY", stormData$evtype, ignore.case = TRUE)] <- "COLD"
stormData$evtype[grep("WARM", stormData$evtype, ignore.case = TRUE)] <- "HEAT"
stormData$evtype[grep("LIGHTNING", stormData$evtype, ignore.case = TRUE)] <- "LIGHTNING"
stormData$evtype[grep("ICY", stormData$evtype, ignore.case = TRUE)] <- "ICE"
stormData$evtype[grep("DAM", stormData$evtype, ignore.case = TRUE)] <- "DAM FAILURE"
stormData$evtype[grep("EROSION", stormData$evtype, ignore.case = TRUE)] <- "EROSION"
stormData$evtype[grep("SLIDE", stormData$evtype, ignore.case = TRUE)] <- "MUD/ROCK SLIDE"
stormData$evtype[grep("RISING WATER", stormData$evtype, ignore.case = TRUE)] <- "HIGH WATER"
stormData$evtype[grep("HOT", stormData$evtype, ignore.case = TRUE)] <- "HEAT"
stormData$evtype[grep("SHOWERS", stormData$evtype, ignore.case = TRUE)] <- "RAIN"
stormData$evtype[grep("OTHER", stormData$evtype, ignore.case = TRUE)] <- "OTHER"
stormData$evtype[grep("FROST", stormData$evtype, ignore.case = TRUE)] <- "FROST"
stormData$evtype[grep("VOLCAN", stormData$evtype, ignore.case = TRUE)] <- "VOLCANO"
stormData$evtype <- as.factor(stormData$evtype)
```

# Results
Here we present the main analytical findings.
## Events Most Harmful with Respect to Population Health

```{r results_health, cache=TRUE, fig.width=14}
healthIssues = stormData %>% group_by(evtype) %>% summarise( fatalities = sum(fatalities), injuries = sum(injuries), totalHarm = injuries + fatalities)
healthIssues = arrange(healthIssues, desc(totalHarm))
top10Healthissues <- head(healthIssues, 10)

#Plot the results
top10Healthissues$evtype <- factor(top10Healthissues$evtype, as.character(top10Healthissues$evtype))
options(scipen=5)
ggplot(top10Healthissues, aes(evtype, totalHarm, fill=evtype)) +
  geom_bar(stat = "identity") +
  ggtitle("Most harmful weather events, 1995-2011") +
  xlab("Event Type") +
  ylab("Total Harm (injuries and fatalities)")
```
## Events with the Greatest Economic Consequences

```{r results_economy, cache=TRUE, fig.width=14}
econ = select(stormData, -fatalities, -injuries)
multiplier = c("K" = 1000, "M" = 1000^2, "B" = 1000^3)
econ$propdmgexp = toupper(econ$propdmgexp)
econ$cropdmgexp = toupper(econ$cropdmgexp)
econ = econ %>% mutate(pdmg = propdmg * multiplier[as.character(propdmgexp)]) %>% mutate(cdmg = cropdmg * multiplier[as.character(econ$cropdmgexp)])
econ[is.na(econ)] = 0
econ = mutate(econ, totalDmg = pdmg + cropdmg)
econ = econ %>% group_by(EVTYPE) %>% summarise(totalDmg = sum(totalDmg))
econ = arrange(econ, desc(totalDmg))
top10EconIssues <- head(econ, 10)

#Plot the results
top10EconIssues$evtype <- factor(top10EconIssues$evtype, as.character(top10EconIssues$evtype))
options(scipen=5)
ggplot(top10EconIssues, aes(evtype, totalDmg, fill=evtype)) +
  geom_bar(stat = "identity") +
  ggtitle("Weather events with greates economic consequences, 1995-2011") +
  xlab("Event Type") +
  ylab("Total Damage (property and crops)")
```